% !TEX root
%
% Copyright (c) 2018  UAVCAN Development Team
% Authors:
%   Pavel Kirienko <pavel.kirienko@zubax.com>
%

\documentclass{uavcandoc}

% To be removed when ready for publication
\usepackage{draftwatermark}
\SetWatermarkScale{1.2}
\SetWatermarkLightness{0.9}

\title{Specification v1.0}

\hbadness=10000

\begin{document}
\frontmatter

\begin{titlepage}

\section*{Overview}

UAVCAN is an open lightweight protocol designed for reliable communication in aerospace and robotic applications via CAN bus.

Features:

\begin{itemize}
    \item Democratic network - no bus master, no single point of failure.
    \item Publish/subscribe and request/response (RPC\footnote{Remote procedure call}) exchange semantics.
    \item Efficient exchange of large data structures with automatic decomposition and reassembly.
    \item Lightweight, deterministic, easy to implement, and easy to validate.
    \item Suitable for deeply embedded, resource constrained, hard real-time systems.
    \item Doubly- or triply- redundant CAN bus support.
    \item Supports high-precision network-wide time synchronization.
    \item The specification and high quality reference implementations in popular programming languages are free,
    open source, and available for commercial use (MIT license).
\end{itemize}

\BeginRightColumn

\section*{Support and feedback}

Information, documentation, and discussions related to UAVCAN are available via the official website at
\href{http://uavcan.org}{uavcan.org}.

\section*{Legal statement}

UAVCAN is an interface standard open to everyone.
No licensing or approval of any kind is necessary for its implementation, distribution, or use.

In no event shall the authors of the standard be liable for any damage arising, directly or indirectly, from its use.

This UAVCAN specification is distributed under the terms of the Creative Commons Attribution 4.0 International License.

\end{titlepage}

\tableofcontents
\BeginRightColumn
\listoffigures
\BeginRightColumn
\listoftables

\mainmatter

\input{introduction/introduction.tex}
\input{basic_concepts/basic_concepts.tex}
\input{dsdl/dsdl.tex}
\input{can_bus_transport_layer/can_bus_transport_layer.tex}
\input{application_layer/application_layer.tex}
\input{hardware_design_recommendations/hardware_design_recommendations.tex}

\end{document}
