\chapter{Application layer}\label{sec:application_layer}

Previous chapters of this specification define a set of basic concepts that are the foundation of the protocol:
they allow one to define data types and exchange data objects over the bus in a robust and deterministic manner.
This chapter is focused on higher-level concepts: rules, conventions, and standard functions that are to be
respected by applications utilizing UAVCAN to maximize cross-vendor compatibility, avoid ambiguities, and
prevent some common design pitfalls.

The rules, conventions, and standard functions defined in this chapter are designed to be an acceptable middle
ground for any sensible aerospace or robotic system.
UAVCAN does not attempt to favor any particular domain or kind of systems among targeted applications.

\begin{itemize}
    \item Section \ref{sec:application_level_requirements} contains a set of mandatory rules that must be
    followed by all UAVCAN implementations.

    \item Section \ref{sec:application_level_conventions} contains a set of conventions and recommendations that
    are not mandatory to follow. Every deviation, however, should be justified and well-documented.

    \item Section \ref{sec:application_level_functions} contains a full list of high-level functions defined on
    top of UAVCAN. Formal specification of such functions is provided in the DSDL data type definition files that
    those functions are based upon (see chapter \ref{sec:list_of_standard_data_types}).
\end{itemize}

\clearpage\input{application_layer/requirements.tex}
\clearpage\input{application_layer/conventions.tex}
\clearpage\input{application_layer/functions.tex}
