\chapter{Transport layer}\label{sec:transport_layer}

This chapter defines the transport layer of UAVCAN.
First, general implementation-agnostic concepts are introduced.
Afterwards, they are further defined for each supported transport medium, e.g., CAN FD.

\section{The concept of transfer}

\subsection{Fundamentals}

A \emph{transfer} is an act of data transmission between nodes.

\subsubsection{Broadcast and unicast transfers}

A transfer that is addressed to any interested node except the source node is a \emph{broadcast transfer}.
A transfer that is addressed to one particular node is a \emph{unicast transfer}.

In the case of broadcast transfers, the sending node makes the data widely available on the bus,
allowing any interested node to freely opt-in and process
it\footnote{The word ``broadcast'' should not lead one to believe that every node is required to
process such transfers. The opt-in logic is facilitated by automatic acceptance filtering features
implemented on the transport layer.}.
The decision of whether to process any given transfer or not is made by receiving nodes.

In the case of unicast transfers, the addressing logic is inverted:
the sending node decides which particular remote node should receive the transfer.
All other nodes remain unaffected by such transmission and take no part in the addressing process.

\subsubsection{Message and service transfers}

A \emph{message transfer} is a broadcast transfer that contains a serialized message and its
metadata\footnote{Such as the subject ID and the source node ID.}.

A \emph{service transfer} is a unicast transfer that contains either a service request or a service response
with related metadata.

\subsubsection{Single-frame and multi-frame transfers}

Both message and service transfers can be further distinguished between single-frame and multi-frame transfers.

A \emph{single-frame transfer} is a transfer that is entirely contained in a single transport frame.
The amount of data that can be exchanged using single-frame transfers is dependent on the transport protocol in use.

A \emph{multi-frame transfer} is a transfer that has its payload distributed over multiple transport frames.
The UAVCAN protocol stack handles transfer decomposition and reassembly automatically.

\subsubsection{Common properties}

The properties listed in the table \ref{table:common_transfer_properties} are common to all types of transfers.

\begin{UAVCANSimpleTable}{Common transfer properties}{|l X|}\label{table:common_transfer_properties}
    Property        & Description \\
    Payload         & The serialized data structure. \\
    Port ID         & A numerical identifier that indicates how the data should be processed.
                      This is the subject ID for message transfers and service ID for service transfers. \\
    Source node ID  & The node ID of the transmitting node (excepting anonymous message transfers). \\
    Priority        & A non-negative integer value that defines the transfer urgency.
                      Higher priority transfers can preempt lower priority transfers. \\
    Transfer ID     & A small overflowing integer that increments with every transfer
                      of this data type from a given node. \\
\end{UAVCANSimpleTable}

\subsection{Message publication}

Message publication is the main method of communication between UAVCAN nodes.

A published message is carried by a single message transfer that contains the serialized message data structure.
A published message does not contain any additional fields besides those listed in the table
\ref{table:common_transfer_properties}.

In order to publish a message, the publishing node must have a node ID that is unique within the network.
An exception applies to \emph{anonymous message publications}.

\subsubsection{Anonymous message publication}

An anonymous message transfer is a transfer that can be sent from a node that does not have a node ID.
This kind of message transfer is especially useful for facilitation of \emph{plug-and-play nodes}
(a high-level concept that is reviewed in detail in chapter \ref{sec:application_layer}).

A node that does not have a node ID is said to be in \emph{passive mode}.
Passive nodes are unable to initiate regular data exchanges,
but they can listen to the data exchanged over the bus,
and they can emit anonymous message transfers.

An anonymous message has the same properties as a regular message, except for the source node ID,
which in the case of anonymous message transfers is always assumed to be zero.

An anonymous transfer can only be a single-frame transfer. Multi-frame anonymous message transfers are not allowed.
This restriction must be kept in mind when designing message data types
intended for use with anonymous message transfers:
when used with anonymous transfers, the whole message must fit into a single transport frame;
however, the same data type can be used with multi-frame regular (non-anonymous) transfers, if desired.

The set of message subject ID values that can be used with anonymous messages may be limited
depending on the transport protocol in use\footnote{This is considered to be an acceptable limitation because
anonymous transfers are intended for an extremely limited set of use cases.}.
It is guaranteed, however, that the subject ID values in the range
from 65528 ($\text{FFF8}_\text{16}$) to 65535 ($\text{FFFF}_\text{16}$), inclusively,
are always available for use with anonymous messages regardless of the transport in use.

Anonymous messages may require special handling logic depending on the transport layer in use.

\subsubsection{Message timing requirements}

Generally, a message transmission should be aborted if it cannot be completed in 1 second.
Applications are allowed to deviate from this recommendation,
provided that every such deviation is explicitly documented.
It is expected that high-frequency high-priority messages may opt for lower timeout values,
whereas low-priority delayable data may opt for higher timeout values to account for the network congestion.

\subsection{Service invocation}

A service invocation sequence consists of two related service transfers:
\emph{service request transfer} and \emph{service response transfer}.

A service request transfer is sent from the invoking node -- \emph{client node} -- to the node
that provides the service -- \emph{server node}.
Upon handling the request, the server node responds to the client node with a service response transfer.
The client will match the response with the corresponding request by comparing the following values:
server node ID, service ID, and the transfer ID.

The tables \ref{table:service_request_transfer_properties} and \ref{table:service_response_transfer_properties}
describe the properties of service request and service response transfers, respectively.

Both the client and the server must have node ID values that are unique within the network;
service invocation is not available to passive nodes.
The client and the server must be two distinct nodes.

\begin{UAVCANSimpleTable}{Service request transfer properties}{|l X|}\label{table:service_request_transfer_properties}
    Property                        & Description \\
    Payload                         & The serialized service request data structure. \\
    Service ID                      & See the table \ref{table:common_transfer_properties}. \\
    Source node ID                  & The node ID of the client (the invoking node). \\
    Destination node ID             & The node ID of the server (the invoked node). \\
    Priority                        & See the table \ref{table:common_transfer_properties}. \\
    Transfer ID                     & An integer value that:
        \begin{enumerate}
            \item allows the server to distinguish the request from other requests from the same client;
            \item allows the client to match the response with its request.
        \end{enumerate} \\
\end{UAVCANSimpleTable}

\begin{UAVCANSimpleTable}{Service response transfer properties}{|l X|}\label{table:service_response_transfer_properties}
    Property                        & Description \\
    Payload                         & The serialized service response data structure. \\
    Service ID                      & Same value as in the request transfer. \\
    Source node ID                  & The node ID of the server (the invoked node). \\
    Destination node ID             & The node ID of the client (the invoking node). \\
    Priority                        & Same value as in the request transfer. \\
    Transfer ID                     & Same value as in the request transfer. \\
\end{UAVCANSimpleTable}

\subsubsection{Service timing requirements}

Applications are recommended to follow the service invocation timing recommendations specified below.
Applications are allowed to deviate from these recommendations,
provided that every such deviation is explicitly documented.

\begin{itemize}
    \item Service transfer transmission should be aborted if does not complete in 1 second.
    \item The client should stop waiting for a response from the server if one has not arrived within 1 second.
\end{itemize}

If the server uses a significant part of the timeout period to process the request,
the client might drop the request before receiving the response.
It is recommended to ensure that the server will be able to process any request in less than 0.5 seconds.

\subsection{Transfer prioritization}\label{sec:transfer_prioritization}

UAVCAN transfers are prioritized by means of the transfer priority property,
which allows at least eight different priority levels for all types of transfers.
Transfers with higher priority levels preempt transfers with lower priority levels,
delaying their transmission until there are no more higher priority transfers to exchange.

The priority level mnemonics and their usage recommendations are specified in the following list.
The mapping between the mnemonics and actual numeric identifiers is transport-dependent.

% https://forum.uavcan.org/t/transfer-priority-level-mnemonics/218/6?u=pavel.kirienko
\begin{description}
    \item[Exceptional] -- The bus designer can ignore these messages when calculating bus load since they
    should only be sent when a total system failure has occurred.
    For example, a self-destruct message on a rocket would use this priority.
    Another analogy is an NMI on a microcontroller.

    \item[Immediate] -- Immediate is a ``high priority message'' but with additional latency constraints.
    Since exceptional messages are not considered when designing a bus, the latency of immediate messages
    can be determined by considering only immediate messages.

    \item[Fast] -- Fast and immediate are both ``high priority messages'' but with additional latency constraints.
    Since exceptional messages are not considered when designing a bus,
    the latency of fast messages can be determined by considering only immediate and fast messages.

    \item[High] -- High priority messages are more important than nominal messages but have looser
    latency requirements than fast messages. This priority is used so that,
    in the presence of rogue nominal messages, important commands can be received.
    For example, one might envision a failure mode where a temperature sensor starts to
    load a vehicle bus with nominal messages.
    The vehicle remains operational (for a time) because the controller is exchanging fast and
    immediate messages with sensors and actuators.
    A system safety monitor is able to detect the distressed bus and command the vehicle to a
    safe state by sending high priority messages to the controller.

    \item[Nominal] -- This is what all messages should use by default.
    Specifically the heartbeat messages should use this priority.

    \item[Low] -- Low priority messages are expected to be sent on a bus under all conditions but cannot
    prevent the delivery of nominal messages.
    They are allowed to be delayed but latency should be constrained by the bus designer.

    \item[Slow] -- Slow messages are low priority messages that have no time sensitivity at all.
    The bus designer need only ensure that, for all possible system states,
    these messages will eventually be delivered.

    \item[Optional] -- These messages might never be sent (theoretically) for some possible system states.
    The system must tolerate never exchanging optional messages in every possible state.
    The bus designer can ignore these messages when calculating bus load.
    This should be the priority used for diagnostic or debug messages that are not required on an operational system.
\end{description}

\subsection{Transfer descriptor}\label{sec:transfer_descriptor}

Transfer emission and reception processes rely on the concept of \emph{transfer descriptor}.

A transfer descriptor is a set of properties that identify a particular set of transfers that originate
from the same source node, share the same port ID, same kind (message or service), and are addressed to the same
destination node (the latter applies only to unicast transfers).

The properties that constitute a transfer descriptor are listed below:

\begin{itemize}
    \item Transfer kind (message or service).
    \item Port ID (subject ID for message transfers, service ID for service transfers).
    \item Source node ID.
    \item Destination node ID (only for service transfers).
\end{itemize}

For convenience, two derived definitions are introduced.
Their objective is to simplify the description of transfer reception and emission logic that appears later in this
specification.
\begin{description}
    \item[Emitted transfer descriptor] -- a transfer descriptor where the source node ID equals the local node's ID.
    \item[Received transfer descriptor] -- a transfer descriptor where the destination node ID equals
    the local node's ID (for service transfers) or is not defined (for message transfers).
\end{description}

\subsubsection{Hard real-time considerations}

Hard real-time applications require a predictable and deterministic data processing time.
The concept of transfer descriptor plays an important role in communication;
hence, its contribution to the worst case data processing load must be carefully analyzed.

From the above definition of transfer descriptor it is easy to derive that for any
message subject ID or any service subject ID the maximum number of transfer descriptors
that can be observed by the local node will never exceed the number of nodes on the bus minus
one\footnote{The local node cannot exchange data with itself, hence minus one.}.
If the number of nodes on the bus cannot be known in advance, it can be considered to equal 127,
which is the maximum node capacity of a UAVCAN bus.

The total number of distinct transfer descriptors that can be observed by a node on any valid UAVCAN bus
is a product of the number of distinct port ID values utilized by the node and the number of other nodes on the bus.

The transport emission and reception logic defined later in this specification relies on data structures
indexed by transfer descriptor values.
Elements of such structures can be easily accessed via constant-complexity static look-up tables
because the worst case number of elements is always statically known.

\section{Transfer emission}

\subsection{Transfer ID computation}\label{sec:transfer_id}

The \emph{transfer ID} is a small unsigned integer value in the range from 0 to 31, inclusive,
that is provided for every transfer.
This value is crucial for many aspects of UAVCAN communication\footnote{One might be tempted to use the transfer ID
value for temporal synchronization of parallel message streams originating from the same node,
where messages bearing the same transfer ID value are supposed to correspond to the same moment of time.
Such use is strongly discouraged because it is impossible to detect if one node is more than
32 messages behind another.
If temporal synchronization is necessary, explicit time stamping should be used instead.};
specifically:
\begin{description}
    \item[Message sequence monitoring] - the continuously increasing transfer ID allows receiving nodes to
    detect lost messages and detect when a message stream from any remote node is interrupted.

    \item[Service response matching] - when a server responds to a request, it uses the same transfer ID for the
    response as in the request,
    allowing any node to emit concurrent requests to the same server while being able to
    match each response with the corresponding request.

    \item[Transport frame deduplication] - for single-frame transfers,
    the transfer ID allows receiving nodes to work around the transport
    frame duplication problem\footnote{This is a well-known issue that can be observed with certain
    transports such as CAN bus -- a frame that appears valid to the receiver may under certain
    (rare) conditions appear invalid to the transmitter, triggering the latter to retransmit the frame,
    in which case it will be duplicated on the side of the receiver.
    Sequence counting mechanisms such as the transfer ID or the toggle bit (both of which are used in UAVCAN)
    allow applications to circumvent this problem.} (multi-frame transfers combat the frame duplication
    problem using the toggle bit).

    \item[Multi-frame transfer reassembly] - more info is provided in section \ref{sec:transfer_reception}.

    \item[Automatic management of redundant interfaces] - the transfer ID parameter allows the UAVCAN protocol
    stack to perform automatic switchover to a back-up interface shall the primary interface fail.
    The switchover logic can be completely transparent to the application, joining several independent
    redundant physical transports into a highly reliable single virtual communication channel.
\end{description}

For message transfers and service request transfers the ID value should be computed as described below.
For service response transfers this value must be directly copied from the corresponding service request transfer.

Every non-passive node must maintain a mapping (or a similar functionally equivalent static
structure\footnote{For example, simple static variables.})
from emitted transfer descriptors (section \ref{sec:transfer_descriptor}) to transfer ID counters.
This mapping is referred to as the \emph{emitted transfer ID map}.

Whenever a node needs to emit a transfer, it will query its transfer ID map for the appropriate transfer descriptor.
If the map does not contain such entry, a new entry will be created with the transfer ID counter initialized to zero.
The node will use the current value of the transfer ID from the map for the transfer,
and then the value stored in the map will be incremented by one.
When the stored transfer ID exceeds its maximum value, it will roll over to zero.

It is expected that some nodes will need to emit certain transfers aperiodically or on an ad-hoc basis,
thereby creating unused entries in the emitted transfer ID map.
In order to avoid keeping unused entries in the map, the nodes are allowed, but not required,
to remove entries from the map that were not used for more than 1 (one) minute
(such behavior is not recommended for hard real-time applications, where constant-complexity deterministic
static structures should be preferred instead).
Therefore, it is possible that a node may emit a transfer with an out-of-order transfer ID value,
if the previous transfer of the same descriptor was emitted more than 1 minute ago.

\subsection{Single frame transfers}

If the size of the entire transfer payload does not exceed the space available for payload in a single transport frame,
the whole transfer will be contained in one transport frame.
Such transfer is called a \emph{single-frame transfer}.

Single frame transfers are more efficient than multi-frame transfers in terms of throughput, latency,
and data overhead.

\subsection{Multi-frame transfers}

\emph{Multi-frame transfers} are used when the size of the transfer payload exceeds the space available
for payload in a single transport frame.

Two new concepts are introduced in the context of multi-frame transfers, both of which are reviewed below in detail:
\begin{samepage}
\begin{itemize}
    \item Transfer CRC\footnote{CRC stands for ``cyclic redundancy check'', an error-detecting code
    added to data transmissions to reduce the likelihood of undetected data corruption.}.
    \item Toggle bit.
\end{itemize}
\end{samepage}

In order to emit a multi-frame transfer, the node must first compute the CRC for the entirety of the transfer payload.
The node appends the resulting CRC value at the end of the transfer payload in the big-endian byte order,
and then emits the resulting byte set in chunks as an ordered sequence of transport frames,
where the first transport frame contains the beginning of the payload bytes,
and the last transport frame contains the last bytes of the payload (possibly none) plus the transfer CRC.

The data field of all transport frames of a multi-frame transfer, except the last one, should be fully utilized.
Applications are allowed to limit the maximum amount of data transferred per transport frame in order to
improve the preemption granularity, thus reducing the worst case latency of higher priority
transfers\footnote{For example, some CAN FD applications may choose to restrict the maximum payload size to 32 bytes
rather than the protocol limit of 64 bytes, as that provides more opportunities for higher-priority frames to
take over the bus. The trade-off is that smaller frames lead to higher transfer fragmentation, increase the bus load,
and increase the overall average latency.}.
Receiving nodes must be prepared to reconstruct multi-frame transfers that utilize the
available payload space partially.

All frames of a multi-frame transfer should be pushed to the transmission queue at once,
in the proper order from the first frame to the last frame.
Explicit gap time between transport frames belonging to the same transfer should not be introduced;
rather, implementations always should strive to minimize it.
Re-ordering of frames belonging to the same multi-frame transfer is prohibited.

\subsubsection{Transfer CRC}\label{sec:transfer_crc}

Transfer CRC allows receiving nodes to ensure that a received multi-frame transfer has been reassembled correctly.

It should be understood that the transfer CRC is not intended for bit-level data integrity checks,
as that must be managed by the transport layer implementation on a per-frame
basis\footnote{A possibility of bit-level errors at the transport frame level may compromise the error-detecting
properties of the transfer CRC.}.
As such, the transfer CRC allows receiving nodes to ensure that all of the frames of a multi-frame
transfer were received, all of the received frames were reassembled in the correct order,
and that all of the received frames belong to the same multi-frame transfer.

The transfer CRC is computed over the entire payload of the transfer.
Certain transport implementations\footnote{Such as CAN FD.} may require a short sequence of padding bytes
to be added at the end of the transfer payload due to the low granularity of the frame payload length property;
in that case, the padding bytes are not to be included in the CRC computation.

The resulting CRC value is appended to the transfer in the \emph{big-endian byte order}
(most significant byte first),
in order to take advantage of the CRC residue check intrinsic to the used algorithm.

The transfer CRC algorithm specification is provided in the table \ref{table:transfer_crc_params}.

\begin{minipage}{0.7\textwidth}
\begin{UAVCANSimpleTable}{Transfer CRC algorithm parameters}{|ll|}\label{table:transfer_crc_params}
    Property        & Value \\
    Name            & CRC-16/CCITT-FALSE \\
    Initial value   & \texttt{0xFFFF} \\
    Polynomial      & \texttt{0x1021} \\
    Reverse         & No \\
    Output XOR      & $0$ \\
    Residue         & $0$ \\
    Check           & $\left(49, 50, \ldots, 56, 57\right) \rightarrow \mathtt{0x29B1}$ \\
\end{UAVCANSimpleTable}
\end{minipage}

The following code snippet provides a basic implementation of the transfer CRC algorithm in C++.

\begin{minipage}{0.9\textwidth}
\begin{minted}{cpp}
// UAVCAN transfer CRC algorithm implementation in C++.
// License: CC0, no copyright reserved.

#include <iostream>
#include <cstdint>
#include <cstddef>

class TransferCRC
{
    std::uint16_t value_ = 0xFFFFU;

public:
    void add(std::uint8_t byte)
    {
        value_ ^= static_cast<std::uint16_t>(byte) << 8U;
        for (std::uint8_t bit = 8; bit > 0; --bit)
        {
            if ((value_ & 0x8000U) != 0)
            {
                value_ = (value_ << 1U) ^ 0x1021U;
            }
            else
            {
                value_ = value_ << 1U;
            }
        }
    }

    void add(const std::uint8_t* bytes, std::size_t length)
    {
        while (length-- > 0)
        {
            add(*bytes++);
        }
    }

    [[nodiscard]] std::uint16_t get() const { return value_; }
};

int main()
{
    TransferCRC crc;
    crc.add(reinterpret_cast<const std::uint8_t*>("123456789"), 9);
    std::cout << std::hex << "0x" << crc.get() << std::endl;  // Outputs 0x29B1
    return 0;
}
\end{minted}
\end{minipage}

\subsubsection{Toggle bit}\label{sec:toggle_bit}

The toggle bit is a property defined at the transport frame level.
Its purpose is to detect and avoid transport frame duplication errors in multi-frame
transfers\footnote{In single-frame transfers, transport frame deduplication is based on the transfer ID counter.}.

The toggle bit of the first transport frame of a multi-frame transfer must be set to one.
The toggle bits of the following transport frames of the transfer must alternate,
i.e., the toggle bit of the second transport frame must be zero,
the toggle bit of the third transport frame must be one, and so on.

For single-frame transfers, the toggle bit must be set to one or removed completely,
whichever option works best for the particular transport.

Transfers where the initial value of the toggle bit is zero must be ignored.
The initial state of the toggle bit may be inverted in the future revisions of the protocol
to facilitate automatic protocol version detection.

\subsection{Redundant interface support}

In configurations with redundant bus interfaces,
nodes are required to submit every outgoing transfer to the transmission queues of
all available redundant interfaces simultaneously.
It is recognized that perfectly simultaneous transmission may not be possible due to different
utilization rates of the redundant interfaces and different phasing of their traffic;
however, that is not an issue for UAVCAN.
If perfectly simultaneous frame submission is not possible, interfaces with lower numerical index
should be handled in the first order.

An exception to the above rule applies if the payload of the transfer depends on some properties
of the interface through which the transfer is emitted.
An example of such a special case is the time synchronization algorithm leveraged by UAVCAN
(documented in chapter \ref{sec:application_layer} of the specification).

Redundant interfaces are used for increased fault tolerance, not for load sharing reasons.
Whenever a node is connected to an interface the likelihood of the interface failing is increased.
This suggests that backup interfaces may only interconnect with mission-critical equipment,
unless a homogeneous network architecture is desired\footnote{Heterogeneous transport configuration
complicates the analysis of the network, which might make it impractical in safety-critical deployments.
In that case, a simpler configuration where each available redundant bus is connected to every node may be
preferred.}.
See section \ref{sec:phy_non_uniform_transport_redundancy}.

\section{Transfer reception}\label{sec:transfer_reception}

\subsection{Transfer ID comparison}

The following explanation relies on the concept of the \emph{transfer ID forward distance}.
Transfer ID forward distance $F$ is a function of two transfer ID values,
$A$ and $B$, that defines the number of increment operations that need to be applied to
$A$ so that $A^\prime{} = B$, assuming modulo 32 arithmetic:
$$A + F = B \quad (\bmod{}\ 32)$$
For example:
$A=0, B=0, F\rightarrow0$;
$A=0, B=5, F\rightarrow5$;
$A=5, B=0, F\rightarrow27$;
$A=31, B=30, F\rightarrow31$;
$A=31, B=0, F\rightarrow1$.

The \emph{half range} of transfer ID is 16.

The following code sample provides an example implementation of the transfer ID comparison algorithm in C++.

\begin{minipage}{0.9\textwidth}  % Mini page is needed to prevent page breaks within the snippet
\begin{minted}{cpp}
// UAVCAN transfer ID forward distance computation algorithm implemented in C++.
// License: CC0, no copyright reserved.

#include <cstdint>
#include <iostream>
#include <cassert>

constexpr std::uint8_t TransferIDBitLength = 5;  // Defined by the specification

[[nodiscard]]
constexpr std::uint8_t computeForwardDistance(std::uint8_t a, std::uint8_t b)
{
    constexpr std::uint8_t MaxValue = (1U << TransferIDBitLength) - 1U;
    assert((a <= MaxValue) && (b <= MaxValue));

    std::int16_t d = static_cast<std::int16_t>(b) - static_cast<std::int16_t>(a);
    if (d < 0)
    {
        d += 1U << TransferIDBitLength;
    }

    assert(d >= 0);
    assert(d <= MaxValue);
    assert(((a + d) & MaxValue) == b);
    return static_cast<std::uint8_t>(d);
}

int main()
{
    assert(0  == computeForwardDistance(0, 0));
    assert(1  == computeForwardDistance(0, 1));
    assert(7  == computeForwardDistance(0, 7));
    assert(0  == computeForwardDistance(7, 7));
    assert(31 == computeForwardDistance(31, 30)); // overflow
    assert(1  == computeForwardDistance(31, 0));  // overflow
    return 0;
}
\end{minted}
\end{minipage}

\subsection{State variables}

\subsubsection{Main principles}

Nodes that receive transfers must keep a certain set of state variables for each
received transfer descriptor (section \ref{sec:transfer_descriptor}).

The set of state variables as documented in the table \ref{table:transfer_receiver_state_variables}
will be referred to as the \emph{receiver state}.
For the purposes of this specification, it is assumed that the node will maintain a
mapping from transfer descriptors to receiver states, which will be referred to as the \emph{receiver map}.
It is understood that implementations might prefer different architectures, which is permitted as
long as the resulting behavior of the node observable at the protocol level is functionally equivalent.

Whenever a node receives a transfer, it will query its receiver map for the matching received transfer descriptor.
If the matching state does not exist, the node will add a new receiver state to the map
and initialize it as defined in section \ref{sec:transfer_reception_initial_state}.
The node then will proceed with the procedure of \emph{receiver state update},
which is defined in section \ref{sec:transfer_reception_state_update_redundant} for redundant transports
and section \ref{sec:transfer_reception_state_update_non_redundant} for non-redundant transports.

It is expected that some transfers will be aperiodic or ad-hoc,
which implies that the receiver map may over time accumulate receiver states that are no longer used.
Therefore, nodes are allowed, but not required, to remove any receiver state from the receiver map
as soon as the state reaches the \emph{transfer ID timeout condition},
as defined in section \ref{sec:transfer_id_timeout_condition}
(such behavior is not recommended for hard real-time applications, where deterministic static look-up tables
should be preferred instead).

Receiver state can only be modified when a new transport frame of a matching transfer is received.
This guarantee simplifies implementation, as it implies that the receiver states will not
require any periodic background maintenance activities.

\begin{UAVCANSimpleTable}{Transfer reception state variables}{|l X|}
    State               & Description \label{table:transfer_receiver_state_variables} \\
    Transfer payload    & Useful payload byte sequence; extended upon reception of new matching transport frames. \\
    Transfer ID         & The transfer ID value of the current transfer or the next expected transfer.
                          Section \ref{sec:transfer_id}. \\
    Next toggle bit     & Expected value of the toggle bit in the next transport frame.
                          Section \ref{sec:toggle_bit}. \\
    Transfer timestamp  & The local monotonic timestamp sampled when the first frame of the transfer arrived.
                          Here, ``monotonic'' means that the reference clock does not change its rate or make leaps. \\
    Interface index     & Only in the case of redundant transport interfaces. \\
\end{UAVCANSimpleTable}

\subsubsection{Initial state}\label{sec:transfer_reception_initial_state}

The initial state is reached when a new entry of the receiver map is created or an existing entry is reset.
Like any other state update, an entry can be created or reset only synchronously with
the reception of a matching transport frame.

Upon reset, the receiver state will meet the following conditions:

\begin{itemize}
    \item The transfer payload buffer is empty.
    \item The transfer ID state matches the actual transfer ID value from the newly received transfer,
    unless this is a non-first frame of a multi-frame transfer.
    In the latter case, the transfer ID state will match the received transfer ID value incremented by one.
    \item The toggle bit is set to its initial state (section \ref{sec:toggle_bit}).
    \item The transfer timestamp matches the reception timestamp from the transport frame.
    \item The interface index matches the index of the interface that the new frame was received from
    (for nodes with redundant interfaces only).
\end{itemize}

\subsubsection{Transfer ID timeout condition}\label{sec:transfer_id_timeout_condition}

A state is said to have reached the transfer ID timeout condition
if the last matching transfer was seen more than 2 (two) seconds ago.
When this condition is reached, the receiver must accept the next transfer disregarding its transfer ID value.

Nodes are allowed to use different timeout values, if that is believed to benefit the application;
however, the chosen timeout value must be strictly lower than the emitted transfer ID map entry timeout
value specified in section \ref{sec:transfer_id} (1 minute).
The ``strictly lower'' requirement is to protect the protocol against possible clock rate discrepancies
on communicating nodes.

If a different timeout value is used, it must be explicitly documented.

Low timeout values increase the risk of undetected transfer duplication when such transfers are significantly
delayed due to bus congestion, which is possible with very low-priority transfers when the bus utilization is high.

High timeout values increase the risk of an undetected transfer loss when a remote node suffers an emitted transfer ID
map state loss (e.g., due to the whole node being restarted).
However, the effects of such a transfer loss caused by a loss of state on a remote node
are always constrained to the first transfer only.

\subsubsection{Interface switchover condition}

This condition is only applicable for configurations with redundant transport interfaces.
It means that the node is allowed to receive the next transfer from an interface that is not the same
the previous transfer was received from.
The condition is reached when the last matching transfer has been successfully received more than
$T_\text{switch}$ seconds ago. The value of $T_\text{switch}$ must not exceed the reception transfer
ID timeout, as defined in section \ref{sec:transfer_id_timeout_condition}.
The actual value of $T_\text{switch}$ can be either a constant chosen by the designer according
to the application requirements (e.g., maximum recovery time in the event of an interface failure),
or the protocol stack can estimate this value automatically by analyzing the transfer intervals.

Nodes are required to let the first interface time out before using the next one because the
transfer ID field is expected to wrap around frequently (every 32 transfers).
Different interfaces are expected to exhibit different latencies even in a properly functioning system,
especially if the system contains both redundantly-interfaced and non-redundantly-interfaced nodes.
If the latency of a backup interface relative to the primary interface exceeds 32 transfer intervals,
and receiving nodes were to be allowed to switch between interfaces freely disregarding the timeout,
the receiving node would skip the whole period of transfer IDs (32 transfers will be lost).
The problem would primarily affect low-priority transfers where large latencies are more likely.

\subsection{State update in a redundant interface configuration}
\label{sec:transfer_reception_state_update_redundant}

The following pseudocode demonstrates the transfer reception process
for a configuration with redundant transport interfaces.

\clearpage
\begin{minted}{cpp}
// Constants:
tid_timeout := 2 seconds;
tid_half_range := 16;
iface_switch_delay := UserDefinedConstant; // Or autodetect

// State variables:
initialized := 0;
payload;
this_transfer_timestamp;
current_transfer_id;
iface_index;
toggle;

function receiveFrame(frame)
{
    // Resolving the state flags:
    tid_timed_out := (frame.timestamp - this_transfer_timestamp) > tid_timeout;
    same_iface := frame.iface_index == iface_index;
    first_frame := frame.start_of_transfer;
    non_wrapped_tid := computeForwardDistance(current_transfer_id, frame.transfer_id) < tid_half_range;
    not_previous_tid := computeForwardDistance(frame.transfer_id, current_transfer_id) > 1;
    iface_switch_allowed := (frame.timestamp - this_transfer_timestamp) > iface_switch_delay;

    // Using the state flags from above, deciding whether we need to reset:
    need_restart :=
        (!initialized) or
        (tid_timed_out) or
        (same_iface and first_frame and not_previous_tid) or
        (iface_switch_allowed and first_frame and non_wrapped_tid);

    if (need_restart)
    {
        initialized := 1;
        iface_index := frame.iface_index;
        current_transfer_id := frame.transfer_id;
        payload.clear();
        toggle := 0;
        if (!first_frame)
        {
            current_transfer_id.increment();
            return;         // Ignore this frame, since the start of the transfer has already been missed
        }
    }

    if (frame.iface_index != iface_index)
    {
        return;  // Wrong interface, ignore
    }

    if (frame.toggle != toggle)
    {
        return;  // Unexpected toggle bit, ignore
    }

    if (frame.transfer_id != current_transfer_id)
    {
        return;  // Unexpected transfer ID, ignore
    }

    if (first_frame)
    {
        this_transfer_timestamp := frame.timestamp;
    }

    toggle := !toggle;
    payload.append(frame.data);

    if (frame.last_frame)
    {
        // CRC validation for multi-frame transfers is intentionally omitted for brevity
        processTransfer(payload, ...);

        current_transfer_id.increment();
        toggle := 0;
        payload.clear();
    }
}
\end{minted}

\subsection{State update in a non-redundant interface configuration}
\label{sec:transfer_reception_state_update_non_redundant}

The following pseudocode demonstrates the transfer reception process for a configuration
with a non-redundant transport interface.
This is a specialization of the more general algorithm defined for redundant transport.

\clearpage
\begin{minted}{cpp}
// Constants:
tid_timeout := 2 seconds;

// State variables:
initialized := 0;
payload;
this_transfer_timestamp;
current_transfer_id;
toggle;

function receiveFrame(frame)
{
    // Resolving the state flags:
    tid_timed_out := (frame.timestamp - this_transfer_timestamp) > tid_timeout;
    first_frame := frame.start_of_transfer;
    not_previous_tid := computeForwardDistance(frame.transfer_id, current_transfer_id) > 1;

    // Using the state flags from above, deciding whether we need to reset:
    need_restart :=
        (!initialized) or
        (tid_timed_out) or
        (first_frame and not_previous_tid);

    if (need_restart)
    {
        initialized := 1;
        current_transfer_id := frame.transfer_id;
        payload.clear();
        toggle := 0;
        if (!first_frame)
        {
            current_transfer_id.increment();
            return; // Ignore this frame, since the start of the transfer has already been missed
        }
    }

    if (frame.toggle != toggle)
    {
        return;  // Unexpected toggle bit, ignore
    }

    if (frame.transfer_id != current_transfer_id)
    {
        return;  // Unexpected transfer ID, ignore
    }

    if (first_frame)
    {
        this_transfer_timestamp := frame.timestamp;
    }

    toggle := !toggle;
    payload.append(frame.data);

    if (frame.last_frame)
    {
        // CRC validation for multi-frame transfers is intentionally omitted for brevity
        processTransfer(payload, ...);

        current_transfer_id.increment();
        toggle := 0;
        payload.clear();
    }
}
\end{minted}

% Please keep \clearpage in front of every transport-specific specification to enforce clear separation!
\clearpage\input{transport_layer/can.tex}
