\section{Serializable types}\label{sec:dsdl_serializable_types}

\subsection{General principles}

Values of the serializable type category can be exchanged between nodes over the Cyphal network.
This is opposed to the expression types (section~\ref{sec:dsdl_expression_types}),
instances of which can only exist while DSDL definitions are being evaluated.
The data serialization rules are defined in section~\ref{sec:dsdl_data_serialization}.

\subsubsection{Alignment and padding}\label{sec:dsdl_serializable_alignment_padding}

For any serializable type,
its \emph{alignment} $A$ is defined as some positive integer number of bits such that the offset of a
serialized representation of an instance of this type relative to the origin of the
containing serialized representation (if any) is an integer multiple of $A$.

Given an instance of type whose alignment is $A$,
it is guaranteed that its serialized representation is always an integer multiple of $A$ bits long.

When constructing a serialized representation,
the alignment and length requirements are satisfied by means of \emph{padding},
which refers to the extension of a bit sequence with zero bits until
the resulting alignment or length requirements are satisfied.
During deserialization, the padding bits are ignored (skipped) irrespective of their value
(also see related section \ref{sec:dsdl_serialization_implicit_truncation}).

\begin{remark}
    For example, given a variable-length entity whose length varies between 1 and 3 bits,
    followed by a field whose type has the alignment requirement of 8,
    one may end up with 5, 6, or 7 bits of padding inserted before the second field at runtime.

    The exact amount of such padding cannot always be determined statically because it depends on the size of the
    preceding entity;
    however, it is guaranteed that it is always strictly less than the alignment requirement of the following field
    or, if this is the last field in a group, the alignment requirement of its container.
\end{remark}

\subsection{Void types}

Void types are used for padding purposes.
The alignment of void types is 1 bit (i.e., no alignment).

Void-typed field attributes are set to zero when an object is serialized and ignored when it is deserialized.
Void types can be used to reserve space in data type definitions for possible use in later versions of the data type.

The DSDL name pattern for void types is as follows: ``\verb|void[1-9]\d*|'',
where the trailing integer represents its width, in bits,
ranging from 1 to 64, inclusive.

Void types can be referred to directly by their name from any namespace.

\subsection{Primitive types}\label{sec:dsdl_primitive_types}

Primitive types are assumed to be known to DSDL processing tools a priori,
and as such, they need not be defined by the user.
Primitive types can be referred to directly by their name from any namespace.

The alignment of primitive types is 1 bit (i.e., no alignment).

\subsubsection{Hierarchy}

The hierarchy of primitive types is documented below.

\begin{itemize}
    \item \textbf{Boolean types.} A boolean-typed value represents a variable of the Boolean algebra.
    A Boolean-typed value can have two values: true and false.
    The corresponding DSDL data type name is ``\verb|bool|''.

    \item \textbf{Algebraic types.} Those are types for which conventional algebraic operators are defined.
    \begin{itemize}
        \item \textbf{Integer types} are used to represent signed and unsigned integer values.
        See table~\ref{table:dsd_integer_properties}.
        \begin{itemize}
            \item \textbf{Signed integer types.} These are used to represent values which can be negative.
            The corresponding DSDL data type name pattern is ``\verb|int[1-9]\d*|'',
            where the trailing integer represents the length of the
            serialized representation of the value, in bits, ranging from 2 to 64, inclusive.

            \item \textbf{Unsigned integer types.} These are used to represent non-negative values.
            The corresponding DSDL data type name pattern is ``\verb|uint[1-9]\d*|'',
            where the trailing integer represents the length of the
            serialized representation of the value, in bits, ranging from 1 to 64, inclusive.
        \end{itemize}

        \item \textbf{Floating point types} are used to approximately represent real values.
        The underlying serialized representation follows the IEEE 754 standard.
        The corresponding DSDL data type name pattern is ``\verb~float(16|32|64)~'', where the trailing
        integer represents the type of the IEEE 754 representation.
        See table~\ref{table:dsd_floating_point_properties}.
    \end{itemize}
\end{itemize}

\begin{CyphalSimpleTable}{Properties of integer types}{|l X l|}%
    \label{table:dsd_integer_properties}%
    Category &
    DSDL names &
    Range, $X$ is bit length \\

    Signed integers &
    \texttt{int2}, \texttt{int3}, \texttt{int4} \ldots{} \texttt{int62}, \texttt{int63}, \texttt{int64} &
    $\left[-\frac{2^{X}}{2},\frac{2^{X}}{2}-1\right]$ \\

    Unsigned integers &
    \texttt{uint1}, \texttt{uint2}, \texttt{uint3} \ldots{} \texttt{uint62}, \texttt{uint63}, \texttt{uint64} &
    $\left[0,2^{X}-1\right]$ \\
\end{CyphalSimpleTable}

\begin{CyphalSimpleTable}{Properties of floating point types}{|X X X X|}
    DSDL name        & Representation    & Approximate epsilon   & Approximate range
    \label{table:dsd_floating_point_properties} \\

    \texttt{float16} & IEEE 754 binary16 & $0.001$               & $\pm{}65504$      \\
    \texttt{float32} & IEEE 754 binary32 & $10^{-7}$             & $\pm{}10^{39}$    \\
    \texttt{float64} & IEEE 754 binary64 & $2 \times{} 10^{-16}$ & $\pm{}10^{308}$   \\
\end{CyphalSimpleTable}

\subsubsection{Cast mode}

The concept of \emph{cast mode} is defined for all primitive types.
The cast mode defines the behavior when a primitive-typed entity is assigned a value that exceeds its range.
Such assignment requires some of the information to be discarded;
due to the loss of information involved, it is called a \emph{lossy assignment}.

The following cast modes are defined:
\begin{description}
    \item[Truncated mode] --- denoted with the keyword ``\verb|truncated|'' placed before the primitive type name.
    \item[Saturated mode] --- denoted with the optional keyword ``\verb|saturated|''
    placed before the primitive type name. If neither cast mode is specified, saturated mode is assumed by default.
    This essentially makes the ``\verb|saturated|'' keyword redundant; it is provided only for consistency.
\end{description}

When a primitive-typed entity is assigned a value that exceeds its range,
the resulting value is chosen according to the lossy assignment rules
specified in table~\ref{table:dsdl_cast_mode}.
Cases that are marked as illegal are not permitted in DSDL definitions.

\begin{CyphalSimpleTable}[wide]{Lossy assignment rules per cast mode}{|l X X|}
    Type category & Truncated mode & Saturated mode (default)
    \label{table:dsdl_cast_mode} \\

    Boolean &
    Illegal: boolean type with truncated cast mode is not allowed. &
    Falsity if the value is zero or false, truth otherwise. \\

    Signed integer &
    Illegal: signed integer types with truncated cast mode are not allowed. &
    Nearest reachable value. \\

    Unsigned integer &
    Most significant bits are discarded. &
    Nearest reachable value. \\

    Floating point &
    Infinity with the same sign, unless the original value is not-a-number, in which case it will be preserved. &
    If the original value is finite, the nearest finite value will be used.
    Otherwise, in the case of infinity or not-a-number, the original value will be preserved. \\
\end{CyphalSimpleTable}

Rules of conversion between values of different type categories do not affect compatibility at the protocol level,
and as such, they are to be implementation-defined.

\subsubsection{Expressions}

At the time of DSDL definition processing,
values of primitive types are represented as instances of the \verb|rational| type (section~\ref{sec:dsdl_rational}),
with the exception of the type \verb|bool|,
instances of which are usable in DSDL expressions as-is.

\begin{CyphalSimpleTable}{Operators defined on instances of type boolean}{|l X X|}
    Op & Type & Description \\

    \texttt{\textbf{!}}     & $(\texttt{bool}) \rightarrow \texttt{bool}$ & Logical not. \\

    \texttt{\textbf{||}}    & $(\texttt{bool}, \texttt{bool}) \rightarrow \texttt{bool}$ & Logical or. \\
    \texttt{\textbf{\&\&}}  & $(\texttt{bool}, \texttt{bool}) \rightarrow \texttt{bool}$ & Logical and. \\

    \texttt{\textbf{==}}    & $(\texttt{bool}, \texttt{bool}) \rightarrow \texttt{bool}$ & Equality. \\
    \texttt{\textbf{!=}}    & $(\texttt{bool}, \texttt{bool}) \rightarrow \texttt{bool}$ & Inequality. \\

\end{CyphalSimpleTable}

\subsubsection{Reference list}

An exhaustive list of all void and primitive types
ordered by bit length is provided below for reference.
Note that the cast mode specifier is omitted intentionally.

\immediate\write18{rm -f ../latex.tmp}
\immediate\write18{../render_list_of_void_and_primitive_types.py > ../latex.tmp}
\immediate\input{../latex.tmp}

\subsection{Array types}

An array type represents an ordered collection of values.
All values in the collection share the same type, which is referred to as \emph{array element type}.
The array element type can be any type except:
\begin{itemize}
    \item void type;
    \item array type\footnote{%
              Meaning that nested arrays are not allowed;
              however, the array element type can be a composite type which in turn may contain arrays.
              In other words, indirect nesting of arrays is permitted.
          }.
\end{itemize}

The number of elements in the array can be specified as a constant expression at the data type definition time,
in which case the array is said to be a \emph{fixed-length array}.
Alternatively, the number of elements can vary between zero and some static limit specified
at the data type definition time,
in which case the array is said to be a \emph{variable-length array}.
Variable-length arrays with unbounded maximum number of elements are not allowed.

Arrays are defined by adding a pair of square brackets after the array element type specification,
where the brackets contain the \emph{array capacity expression}.
The array capacity expression shall yield a positive integer of type ``\verb|rational|'' upon its evaluation;
any other value or type renders the current DSDL definition invalid.

The array capacity expression can be prefixed with the following character sequences in order to define
a variable-length array:
\begin{itemize}
    \item ``\verb|<|'' (ASCII code 60) --- indicates that the integer value yielded by the array capacity expression
    specifies the non-inclusive upper boundary of the number of elements.
    In this case, the integer value yielded by the array capacity expression shall be greater than one.

    \item ``\verb|<=|'' (ASCII code 60 followed by 61) --- same as above, but the upper boundary is inclusive.
\end{itemize}
If neither of the above prefixes are provided, the resultant definition is that of a fixed-length array.

The alignment of an array equals the alignment of its element type\footnote{
    E.g., the alignment of \texttt{uint5[<=3]} or \texttt{int64[3]} is 1 bit (that is, no alignment).
}.

\subsection{Composite types}\label{sec:dsdl_composite_types}

\subsubsection{Kinds}

There are two kinds of composite type definitions: message types and service types.
All versions of a data type shall be of the same kind\footnote{%
    For example, if a data type version 0.1 is of a message kind, all later versions of it shall be messages, too.
}.

A service type defines two inner data types:
one for service request object, and one for service response object, in that order.
The two types are separated by the service response marker (``\verb|---|'') on a separate line.

The presence of the service response marker indicates that the data type definition at hand is of the service kind.
At most one service response marker shall appear in a given definition.

\subsubsection{Dependencies}

In order to refer to a composite type from another composite type definition
(e.g., for nesting or for referring to an external constant),
one has to specify the full data type name of the referred data type followed by its
major and minor version numbers separated by the namespace separator character,
as demonstrated on figure~\ref{fig:dsdl_nested_reference}.

To facilitate look-up of external dependencies,
implementations are expected to obtain from external sources\footnote{%
    For example, from user-provided configuration options.
} the list of directories that are the roots of namespaces containing the referred dependencies.

\begin{figure}[H]
    $$
    \underbrace{\texttt{\huge{uavcan.node.Heartbeat}}}_{\text{full data type name}}%
    \texttt{\huge{.}}%
    \underbrace{\texttt{\huge{1.0}}}_{\substack{\text{version} \\ \text{numbers}}}%
    $$
    \caption{Reference to an external composite data type definition\label{fig:dsdl_nested_reference}}
\end{figure}

If the referred data type and the referring data type share the same full namespace name,
it is allowed to omit the namespace from the referred data type specification
leaving only the short data type name, as demonstrated on figure~\ref{fig:dsdl_nested_reference_short}.
In this case, the referred data type will be looked for in the namespace of the referrer.
Partial omission of namespace components is not permitted.

\begin{figure}[H]
    $$
    \underbrace{\texttt{\huge{Heartbeat}}}_{\text{short data type name}}%
    \texttt{\huge{.}}%
    \underbrace{\texttt{\huge{1.0}}}_{\substack{\text{version} \\ \text{numbers}}}%
    $$
    \caption{Reference to an external composite data type definition located in the same namespace
             \label{fig:dsdl_nested_reference_short}}
\end{figure}

Circular dependencies are not permitted.
A circular dependency renders all of the definitions involved in the dependency loop invalid.

If any of the referred definitions are marked as deprecated,
the referring definition shall be marked deprecated as well\footnote{%
    Deprecation is indicated with the help of a special directive, as explained in section~\ref{sec:dsdl_directives}.
}.
If a non-deprecated definition refers to a deprecated definition,
the referring definition is malformed\footnote{%
    This tainting behavior is designed to prevent unexpected breakage of
    type hierarchies when one of the deprecated dependencies reaches its end of life.
}.

When a data type is referred to from within an expression context,
it constitutes a literal of type ``\verb|metaserializable|'' (section~\ref{sec:dsdl_metaserializable}).
If the referred data type is of the message kind,
its attributes are accessible in the referring expression through application of the
attribute reference operator ``\verb|.|''.
The available attributes and their semantics are documented in the section~\ref{sec:dsdl_local_attributes}.

\begin{remark}
    \begin{minted}{python}
        uint64 MY_CONSTANT = vendor.MessageType.1.0.OTHER_CONSTANT
        uint64 MY_CONSTANT = MessageType.1.0.OTHER_CONSTANT
        # The above is valid if the referring definition and the referred definition
        # are located inside the root namespace "vendor".
        @print MessageType.1.0
    \end{minted}
\end{remark}

\subsubsection{Tagged unions}\label{sec:dsdl_composite_tagged_unions}

Any data type definition can be supplied with a special directive (section~\ref{sec:dsdl_directives})
indicating that it forms a tagged union.

A tagged union type shall contain at least two field attributes.
A tagged union shall not contain padding field attributes.

The value of a tagged union object is a function of the field attribute which value it is currently holding
and the value of the field attribute itself.

To avoid ambiguity, a data type that is not a tagged union is referred to as a \emph{structure}.

\subsubsection{Alignment and cumulative bit length set}\label{sec:dsdl_composite_alignment_cumulative_bls}

The alignment of composite types is one byte (8 bits)\footnote{%
    Regardless of the content.
    It follows that empty composites can be inserted arbitrarily to force byte alignment
    of the next field(s) at runtime.
}.

Per the definitions given in \ref{sec:dsdl_serializable_alignment_padding},
a serialized representation of a composite type is padded to 8 bits by inserting padding bits
after its last element until the resulting length is a multiple of 8 bits.

Given a set of field attributes, their \emph{cumulative bit length set} is computed
by evaluating every permutation of their respective bit length sets plus the required padding.
\begin{itemize}
    \item For tagged unions, this amounts to the union of the bit length sets of each field
    plus the bit length set of the implicit union tag.
    \item Otherwise, the cumulative bit length set is the Cartesian product of the bit length sets of each field
    plus the required inter-field padding.
\end{itemize}
Related specifics are given in section \ref{sec:dsdl_data_serialization} on data serialization.

\subsubsection{Extent and sealing}\label{sec:dsdl_composite_extent_and_sealing}

As detailed in section \ref{sec:dsdl_versioning},
data types may evolve over time to accommodate new design requirements, new features, to rectify issues, etc.
In order to allow gradual migration of deployed systems to revised data types,
it is desirable to ensure that they can be modified in a way that does not render new definitions
incompatible with their earlier versions.
In this context there are two related concepts:

\begin{description}
    \item[Extent] --- the minimum amount of memory, in bits, that shall be allocated to store the serialized
    representation of the type.
    The extent of any type is always greater than or equal the maximal value of its bit length set.
    It is always a multiple of 8.

    \item[Sealing] --- a type that is \emph{sealed} is non-evolvable and its extent equals the maximal value
    of its bit length set\footnote{%
        I.e., the smallest possible extent.
    }.
    A type that is not sealed is also referred to as \emph{delimited}.
\end{description}

The extent is the growth limit for the maximal bit length of the type as it evolves.
The extent should be at least as large as the longest serialized representation of any compatible version of the type,
which ensures that an agent leveraging a particular version can correctly process any other compatible
version due to the avoidance of unintended truncation of serialized representations.

Serialized representations of evolvable definitions may carry additional metadata which introduces a certain overhead.
This may be undesirable in some scenarios, especially in cases where serialized representations of the
definition are expected to be highly compact, thereby making the overhead comparatively more significant.
In such cases, the designer may opt out of the extensibility by declaring the definition sealed.
Serialized representations of sealed definitions do not incur the aforementioned overhead.

The related mechanics are described in section \ref{sec:dsdl_serialization_composite_non_sealed}.

\begin{figure}[H]
    $$
    \overbrace{%
        \underbrace{%
            \blacksquare\blacksquare\blacksquare\blacksquare\blacksquare\blacksquare%
            \blacksquare\blacksquare\blacksquare\blacksquare\blacksquare\blacksquare%
        }_{\substack{\text{Longest serialized} \\ \text{representation}}}%
        \underbrace{%
            \boxtimes\boxtimes\boxtimes\boxtimes\boxtimes\boxtimes\boxtimes\boxtimes%
        }_{\substack{\text{Memory reserve} \\ \text{(none if sealed)}}}%
    }^{\substack{%
        \text{Extent} \\
        \text{(memory requirement)}
    }}
    $$
    \caption{Serialized representation and extent\label{fig:dsdl_extent}}
\end{figure}

\begin{remark}
    Because of Cyphal's commitment to determinism, memory buffer allocation can become an issue.
    When using a flat composite type (where each field is of a primitive type) with the implicit truncation rule,
    it is clear that the last defined fields are to be truncated out shall the allocated buffer be too small
    to accommodate the serialized representation in its entirety.
    If there is a composite-typed field, this behavior can no longer be relied on.
    The technical details explaining this are given in section \ref{sec:dsdl_serialization_composite_non_sealed}.

    Conventional protocols manage this simply by delaying the memory requirement identification until runtime,
    which is unacceptable to Cyphal.
    The solution for Cyphal is to allow the data type author to require implementations to reserve more memory
    than required by the data type definition unless it is \verb|@sealed|
    (or unless the implementation does use dynamic memory allocation).
\end{remark}

The extent shall be set explicitly using the directive described in section \ref{sec:dsdl_directive_extent},
unless the definition is declared sealed using the directive described in section \ref{sec:dsdl_directive_sealed}.
The directives are mutually exclusive.

It is allowed for a sealed composite type to nest non-sealed (delimited) composite types, and vice versa.

\subsubsection{Bit length set}

The bit length set of a sealed composite type equals the cumulative bit length set
of its fields plus the final padding (see section \ref{sec:dsdl_composite_alignment_cumulative_bls}).

\begin{remark}
    The bit length set of the following is $\left\{ 8, 24, 40, 56 \right\}$:
    \begin{minted}{python}
        uint16[<=3] foo
        @sealed
    \end{minted}

    The bit length set of the following is $\left\{ 16, 32, 48, 64 \right\}$:
    \begin{minted}{python}
        uint16[<=3] foo
        int2 bar
        @sealed
    \end{minted}

    The bit length set of the following is $\left\{ 8, 16 \right\}$:
    \begin{minted}{python}
        bool[<=3] foo
        @sealed
    \end{minted}
\end{remark}

The bit length set of a non-sealed (delimited) composite type is dependent only on its extent $X$,
and is defined as follows:
$$
    \left\{ B_\text{DH} + 8b \mid b \in \mathbb{Z},\ 0 \leq b \leq \frac{X}{8} \right\}
$$
where $B_\text{DH}$ is the bit length of the \emph{delimiter header}
as defined in section \ref{sec:dsdl_serialization_composite_non_sealed}.

\begin{remark}
    This is intentionally not dependent on the fields of the composite because the definition of delimited
    composites should be possible to change without violating the backward compatibility.

    If the bit length set was dependent on the field composition, then a composite $A$ that nests another composite
    $B$ could have made a fragile assumption about the offset of the fields that follow $B$
    that could be broken when $B$ is modified. Example:

    \begin{minted}{python}
        # A.1.0
        B.1.0 x
        float32 assume_aligned  # B.1.0 contains a single uint64, assume this field is 32-bit aligned?
        @sealed
    \end{minted}

    \begin{minted}{python}
        # B.1.0
        uint64 x
        @extent 17 * 8
    \end{minted}

    Imagine then that \verb|B.1.0| is replaced with the following:

    \begin{minted}{python}
        # B.1.1
        uint64 x
        bool[<=64] y
        @extent 17 * 8
    \end{minted}

    Once this modification is introduced, the fragile assumption about the alignment of
    \verb|A.1.0.assume_aligned| would be violated.
    To avoid this issue, the bit length set definition of delimited types intentionally discards the information
    about its field composition, forcing dependent types to avoid any non-trivial assumptions.
\end{remark}

When serializing an object, the amount of memory needed for storing its serialized representation
may be taken as the maximal value of its bit length set minus the size of the delimiter header,
since this bound is tighter than the extent yet guaranteed to be sufficient.
This optimization is not applicable to deserialization since the actual type of the object may not be known.

\subsubsection{Type polymorphism and equivalency}

Type polymorphism is supported in DSDL through structural subtyping.
The following definition relies on the concept of \emph{field attribute},
which is introduced in section~\ref{sec:dsdl_attributes}.

Polymorphic relations are not defined on service types.

Let $B$ and $D$ be DSDL types that define $b$ and $d$ field attributes, respectively.
Let each field attribute be assigned a sequential index according to its position in the DSDL definition
(see section~\ref{sec:dsdl_grammar} on statement ordering).

\begin{enumerate}
    % -----------------------------------------------------------------------------------------------------------------
    \item Structure subtyping rule --- $D$ is a \emph{structural subtype} of $B$ if all conditions are satisfied:
    \begin{itemize}
        \item neither $B$ nor $D$ define a tagged union\footnote{%
            This is because tagged unions are serialized differently.
        };
        \item neither $B$ nor $D$ are sealed\footnote{%
            Sealed types are serialized in-place as if their definition was directly copied into the outer
            (containing) type (if any).
            This circumstance effectively renders them non-modifiable because that may break the bit layout
            of the outer types (if any).
            More on this in section \ref{sec:dsdl_serialization_composite_non_sealed}.
        };
        \item the extent of $B$ is not less than the extent of $D$\footnote{%
            This is to uphold the Liskov substitution principle.
            A deserializer expecting an instance of $B$ in a serialized representation should be invariant
            to the replacement $B \leftarrow{} D$.
            If the extent of $D$ was larger, then its serialized representation could spill beyond the allocated
            container, possibly resulting in the truncation of the following data, which in turn could result in
            incorrect deserialization.
            See \ref{sec:dsdl_data_serialization}.
        };
        \item $B$ is not $D$;
        \item $b \leq d$;
        \item for each field attribute of $B$ at index $i$ there is an equal\footnote{%
            Field attribute equality is defined in section~\ref{sec:dsdl_attributes}.
        } field attribute in $D$ at index $i$.
    \end{itemize}

    % -----------------------------------------------------------------------------------------------------------------
    \item Tagged union subtyping rule --- $D$ is a structural subtype of $B$ if all conditions are satisfied:
    \begin{itemize}
        \item both $B$ and $D$ define tagged unions;
        \item neither $B$ nor $D$ are sealed;
        \item the extent of $B$ is not less than the extent of $D$;
        \item $B$ is not $D$;
        \item $b \leq d$;
        \item $2^{\lceil\log_2 \text{max}\left(8, \lceil\log_2 b\rceil\right)\rceil} =
               2^{\lceil\log_2 \text{max}\left(8, \lceil\log_2 d\rceil\right)\rceil}$\footnote{%
            I.e., the length of the implicit union tag field should be the same.
        };
        \item for $i \in \left[0, b\right)$, the type of the field attribute of $D$ at index $i$
        is the same or is a subtype of the type of the field attribute of $B$ at index $i$.
        \item for $i \in \left[0, b\right)$, the name of the field attribute of $D$ at index $i$
        is the same as the name of the field attribute of $B$ at index $i$.
    \end{itemize}

    % -----------------------------------------------------------------------------------------------------------------
    \item Empty type subtyping rule --- $D$ is a structural subtype of $B$ if all conditions are satisfied:
    \begin{itemize}
        \item $b = 0$\footnote{%
            If $B$ contains no field attributes, the applicability of the Liskov substitution principle is invariant to
            whether its subtypes are tagged union types or not.
        };
        \item neither $B$ nor $D$ are sealed;
        \item the extent of $B$ is not less than the extent of $D$;
        \item $B$ is not $D$.
    \end{itemize}

    % -----------------------------------------------------------------------------------------------------------------
    \item Header subtyping rule --- $D$ is a structural subtype of $B$ if all conditions are satisfied:
    \begin{itemize}
        \item neither $B$ nor $D$ define a tagged union;
        \item both $B$ and $D$ are sealed;
        \item the first field attribute of $D$ is of type $B$.
    \end{itemize}
\end{enumerate}

If $D$ is a structural subtype of $B$, then $B$ is a \emph{structural supertype} of $D$.

$D$ and $B$ are \emph{structurally equivalent} if $D$ is a structural subtype and a structural supertype of $B$.

A \emph{type hierarchy} is an ordered set of data types such that for each pair of its members
one type is a subtype of another, and for any member its supertypes are located on the left.

\begin{remark}
    Subtyping example for structure (non-union) types. First type:

    \begin{minted}{python}
        float64 a       # Index 0
        int16[<=9] b    # Index 1
        @extent 32 * 8
    \end{minted}

    The second type is a structural subtype of the first type:

    \begin{minted}{python}
        float64 a       # Index 0
        int16[<=9] b    # Index 1
        uint8 foo       # Index 2
        @extent 32 * 8
    \end{minted}

    Subtyping example for union types. First type:

    \begin{minted}{python}
        @union                            # The implicit union tag field is 8 bits wide
        uavcan.primitive.Empty.1.0 foo
        float16 bar
        uint8 zoo
        @extent 128 * 8
    \end{minted}

    The second type is a structural subtype of the first type:

    \begin{minted}{python}
        @union                            # The implicit union tag field is 8 bits wide
        uavcan.diagnostic.Record.1.0 foo  # Subtype
        float16 bar                       # Same
        uint8 zoo                         # Same
        int64[<=64] baz                   # New field
        @extent 128 * 8
    \end{minted}

    A structure type that defines zero fields is a structural supertype of any other structure type,
    regardless of either or both being a union, provided that its extent is sufficient.
    A structure type may have an arbitrary number of supertypes as long as the field equality constraint is satisfied.

    Header subtyping example. The first type is named \verb|A.1.0|:

    \begin{minted}{python}
        float64 a
        int16[<=9] b
        @sealed
    \end{minted}

    The second type is a structural subtype of the first type:

    \begin{minted}{python}
        A.1.0 base
        uint8 foo
        @sealed
    \end{minted}
\end{remark}

\begin{remark}[breakable]
    The following example in C demonstrates the concept of polymorphic compatibility detached from DSDL.

    \begin{samepage}
        \begin{minted}{c}
            struct base
            {
                int a;
                float b;
            };

            struct derived_first
            {
                int a;
                float b;
                double c;
            };

            struct derived_second
            {
                int a;
                float b;
                short d;
            };

            float compute(struct base* value)
            {
                return (float)value->a + value->b;
            }

            int main()
            {
                struct derived_first foo  = { .a = 123, .b = -456.789F, .c = 123.456 };
                struct derived_second bar = { .a = -123, .b = 456.789F, .d = -123 };
                // Both derived_first and derived_second are structural subtypes of base. The program returns zero.
                return compute(&foo) + compute(&bar);
            }
        \end{minted}
    \end{samepage}
\end{remark}
